### 2009-12-08 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
First Draft of the dissertation peer reviewed with Kleber.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@680 559edde2-a4d4-44e0-96fc-4035e40d7396

\section{Database Systems Contenders}

This section describes the database systems contenders considered being used
together with for the implementation of the Data Persistence component for 
NetBEAMS. As described in the previous chapter, the selection of the
technologies must be based on the list of functional and non-functional
requirements specified within the scope this work, defined in the previous
section.

One of the essential questions related to data persistence is regarding the
data model and nature of the database, that is, the use of the relational or
structured models. In addition to that, and given the present time of academic
research and recent trends distributed computing such as Cloud Computing and
the Internet \cite{cloud-comp-survey}, the selection of a data persistence for
sensor networks should maintain a wider spectrum of the types of database 
systems. Among other intriguing questions, \cite{db-is-rdbs-dommed} reflects
about the use of alternatives to the relational model for dynamic environments
such as data persistence for the Web. Similarly, these trends are also related
to the advances of Cloud Computing and its persistence layer as a solution to
dynamic systems \cite{cloud-comp-architectures}. For this reason, this work
also included the overview of alternative data models for the development of
an experiment that brings data persistence to NetBEAMS. In such a manner, the
list of the systems considered for assessment is the following:

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  dateware house with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks
  implementation \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database
  system described in the article that questions the use of the relational
  model \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. It was selected because NetBEAMS can
  represent the collected data in the XML format as shown in setion 
  \ref{sec:dsp-message}.
\end{itemize}

These technologies will be reviewed in contrast to their capabilities to adhere
to the specifications of the taxonomies defined in chapter 3 and the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the network
  sink at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for for
  distribution using the OPEnDAP format;
  \item An automated approach to the data collection procedures is done using 
  NetBEAMS \cite{netbeams2009}, using software components to obtain and send
  the data to the network sink. A full description of this approach is
  described in section \ref{netbeams-architecture}.
\end{itemize}

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}. After data is
collected from the sensors, it needs to be processed and archived for reuse by
different types of users, ranging from researchers, students, and the public
over the Internet.

@@

As it is supposed to be, all the contenders are developed to support data
persistence and management, what is the requirement for the Data Archival
purpose of the sensor network data.

Basically, all the selected database systems support the purpose of the data.

@@

As shown in the previous section, NetBEAMS's architecture is based on a
single-hop star infrastructure from the SF-BEAMS sensor network, having all the
data collected from the sensors being immediately sent to the network sink
(Section \ref{sec:sn-infrastructure}).

Based on the observations also described in the last chapter, the simplest
position to save the collected data is through the use of an \textbf{External
Storage}. In addition to that, the use of a Data-Centric Storage strategy is
also considered in situations where the data load in the network sink reaches
its storage capacity. The \textbf{Data-Centric Storage} storage separates the
location of storage of the data based in any property of the data such as the
value of a given key. Such approach of data partitioning is provided by
database system in two different flavors.

The Data-Centric storage separates the location of storage of the data based in
any property of the data such as the value of a given key, etc. Such approach
of data partitioning is provided by database system in two different flavors
\cite{db-partitioning-relational}:

\begin{itemize}
  \item \textbf{Horizontal Partitioning}: the data segmentation is done based
  on the values of the table rows in different groups allocated to different
  datasets physically located in different servers. In this way, all the
  columns defined for a given entity are found on each partition. For instance,
  horizontal partitioning can be defined as a collection five years worth of
  historical sensor data being partitioned into five separate servers, where
  each server has a single year's worth of sensor data;
  \item \textbf{Vertical Partitioning}: data is partitioned vertically in a way
  that the columns of an entity are divided into different database tables. In
  this way, the particular dataset is broken down to different tables. For
  example, if the database is used to store pictures of a given sensor camera
  in a BLOB column, it would be the most preferable to the scenario where the
  images are not constantly referenced when compared to the textual data.
\end{itemize}

@@

The External Storage approach is the most common way to implement persistence
for any breed of systems since data management is usually offered by most
database systems out-of-the-box. Furthermore, this approach can be accompanied
with the use of Distributed Systems approaches such as Replication to help a
system to scaling in terms of performance.

@@

The most common problem related to a single External Data Storage is that 
resources may run into data management problems such as full-disk space or
disk failures.

@@

The benefit of the Data-Centric approach can be directly associated with
database partitioning mechanisms described earlier. Since data is distributed
into different locations, this approach decreases the size of datasets when the
database server performs a query processing in a database table. For this
reason, this approach helps the database administrators to scale their database
infrastructure and improve the database performance.

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-table-partition, db-partitioning-relational}. Both former and the
latter are used in regular schema-dependent models or schema-less models,
which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas, db-partitioning-relational}.

\begin{itemize}
  \item \textbf{Directory Based Partitioning}: an ordinary lookup service is
  used to identify which database shard to use, where the lookup table can be
  stored in a secondary area;
  \item \textbf{Vertical Partitioning}: each feature of the system is usually
  represented by a table in the relational model and placed in a different
  server host. For example, each sensor in a given network can be divided by
  its device type;
  \item \textbf{Key or Hash Based Partitioning}: the value of a given key
  property can  be used as the key of the shard, or used as an input to a hash
  function, whose output determines which database shard location the data is
  going to  be stored;
  \item \textbf{Range-Based Partitioning}: after partitioning each feature on
  a different  server host, different multiple shards, or similar hosts, can be
  added by using  the range of a given key of the entity of a function. For
  example, different  sensor data can be distributed by geographic origin of
  the data.
\end{itemize}

An example of a Data-Centric storage with different shards
\cite{db-shard-intro} ca n be see in Figure
\ref{fig:database-sharding-by-region}, where data is segregated by the
location of data. Each shard is assumed to be placed in different server
hosts. Since the data model is denormalized, the organization of data is
entirely placed in each of these shards in a Horizontal partition of the
entity. As it is shown, the shards by location can be defined by a shard key.

@@

The data partitioning is a highly restricted technology, normally available to
advanced users of distributed database systems. Despite being a capability
being offered common-off-the-shelf by some popular database systems, this
strategy requires closer database management. Some of the problems related to
this strategy are regarding its execution \cite{db-shard-discussion}:

\begin{itemize}
  \item \textbf{Rebalancing Shards}: If the schema model changes for the
  shards, the strategy of rebalancing must be done. For instance, when one of
  the shards out-grows more than the others. As a consequence, the database
  shards must be rebalanced by transferring data to new locations whenever
  necessary. This technique is starting to appear as COTS implementations such
  as mongoDB \cite{mongodb} or MySQL \cite{mysql};
  \item \textbf{Referential integrity}: any cross-shard queries can represent
  problems since the application layer is responsible to enforce data
  integrity. Examples include verifying constraints of foreign keys when the
  partition is done using Vertical table partitioning, considering that data
  collections are organized in separate spaces, and physically placed in
  different shard nodes \cite{db-table-partition}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports External or Data-Centric Storage approaches.
  When Data-Centric Storage is used, both vertical and horizontal partitions
  \cite{db-partitioning-relational} can be chosen, using different strategies
  to partition the data;
  \item \textbf{TinyDB}: Only supports the External Storage;
  \item \textbf{MongoDB}: Supports External or Data-Centric Storage. When the
  latter approach is used, the horizontal partition strategy is available with
  automatic shards \cite{db-mongo-partition};
  \item \textbf{DB2}: Supports External and Data-Centric storage. Only supports
  vertical table partition \cite{db-db2-partition} for the data-centric storage.
\end{itemize}

All the selected database systems support the External Storage method. However,
they all differ with the support to Data-Centric Storage support.

\section{Analysis of the Data Model Taxonomy}

One of the most common practices in the area of database system is to use the
relational model to persist data, although the application of the system may
not fit to solve the problem. In fact, the main users of sensor networks may
not hold any expertise in database systems or data modeling, given they come
from different science areas. Taking NetBEAMS as an example, we see a Sensor
Network managed by Marine Biologists without expertise in Data Management,
Modeling systems, and for this reason, one of the requirements of the system is
trying the use of schema-less approaches. The \textbf{Key-Value-Pair Data
Model} or the \textbf{Document-Oriented Data Model} seems to be the simplest
choices of the models.

@@

\begin{itemize}
  \item Constant data schema changes require constant database normalization
  process, changes to structure, database management, etc. Schema-Dependent 
  databases always require refactoring of the schema to accommodate changes to
  the entites properties;
  \item The approach of Relational Model might does not fit the nature of
  collected data from wired or wireless sensor devices, since they tend to
  change over time \cite{db-is-rdbs-dommed}. They usually does not give
  support time-series data nor provenance well \cite{sn-provenance};
  \item Some research have suggested changes to the SQL clauses to better
  support time-series when collecting data using the relational model 
  \cite{sn-db-newop}.
\end{itemize}

Other projects have been using the XML data models for persistence. It falls
into the same category as the Relational Model using XML documents \cite{xml},
being complaint to an XML Schema \cite{xml-schema}. This model uses XPath
\cite{xml-xpath} technology for querying documents, although some hybrid
technologies may still use SQL \cite{sql} for that matter \cite{db2}. In this
way, a materialized version of the data from the defined previous schema can be
seen in Figure \ref{fig:persistence-example-relational}.

@@

Although this data model has been used in different types of applications,
developers have tried to use the Relational Model to model an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. With the emergence of Internet
applications, the necessity of such a model that does not require constant
refactoring led developers to propose such manner using the relational model as
documented in technical articles such as \cite{db-kvp-in-relational01} and
\cite{db-kvp-in-relational02}. Based on these articles, consider the
relational model depicted in Figure \ref{fig:KVP-on-Relational-Model} as a
data model for persisting data to our case study using the key-value strategy.
The problem with it is that key repetition occurs, since the nature of
time-series data requires the use of a timestamp key for each property of the
sensor. Therefore, this strategy is not well-suited to provide persistence
data from collected sensor networks.

@@

The growth of distributed systems and the Internet have enabled the development
of powerful database servers that can be organized in the context of
infrastructure and how to represent models. An powerful up-coming approach is
the use of database systems that do not use structured language for querying
its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, which is also known as Document-Oriented, Distributed Hash
Table. Such a model has the following properties:

@@

There are no records of the use of this data model with Sensor Networks.
Different variants of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure that can be freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json} in the implementation
of database systems which are becoming popular with the new trend called Cloud
Computing \cite{cloud-comp-architectures}. Databases implementing this
strategy is the mongoDB .

@@

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describe the data, depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships;
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

@@

\begin{itemize}
  \item \textbf{MySQL}: supports the Relational Model;
  \item \textbf{TinyDB}: supports the Relational Model; 
  \item \textbf{MongoDB}: supports Document-Oriented Model;
  \item \textbf{IBM DB2}: supports the Relational Model and the XML model.
\end{itemize}

\section{Analysis of the Query Processing Mechanism Taxonomy}

@@

All the database systems technology support a centralized query system.

\section{Analysis of the System Organization Taxonomy}

@@

Each of the databases were scored according to its support to the different
types of taxonomies, and randomly selected by using the literature review's
list, as well as the Internet. Since mongoDB is based on a schema-less database
system, has not been reported in use by the sensor network community, and
supports most of the taxonomies defined in the previous chapter, it was
selected for the implementation of the DSP Data Persistence component for
NetBEAMS.

\begin{table}
    \label{tab:ysi-data-distribution}
    \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{TinyDB} & \textbf{MongoDB} & \textbf{IBM DB2}\\\hline
        Centralized Query & + & + & + & + \\\hline 
        Distributed System & ++ & - & ++ & +\\\hline 
        Schema-less & - & - & + & -\\\hline 
        Provenance Support & + & + & + & +\\\hline 
        NO-SQL Query & - & - & + & -\\\hline 
        Partition & + & - & + & +\\\hline 
        Hot Backup & + & - & + & +\\\hline 
        Export Capability & + & - & + & -\\\hline 
        Programming Lang. & + & + & + & +\\\hline
        Script Lang. & + & - & + & +\\\hline
        Open-Source & + & + & + & -\\\hline
        \end{tabular}
    \end{center}
\end{table}

>>>>

\section{Database System ``Contenders''}

This section describes the database systems ``contenders'' considered to be used
as the persistence technology backend for NetBEAMS. As described in the
previous chapter, the selection of the technologies must be based on the list
of functional and non-functional requirements specified within the scope of
this work.

One of the essential questions related to data persistence is regarding the
data model and breed of the database. That is, the use of the relational or
any other model to describe data along with its management. Traditionally, the
relational data model has been used to provide data persistence for different
types of projects, inclusing sensor networks \cite{sn-dataware-house,
db-xml-enabled, sn-db-tinydb}. However, tradition does not always get translated
to efficiency in terms of problem solving. For this reason,
\cite{db-is-rdbs-dommed} reflects about the use of alternatives to the
relational model when the data persistence problem in consideration is to
provide data persistence for dynamic environments as Internet applications. The
author also suggests that the relational model is hard to maintain in
ever-changing environments, and thus, proposes the use of the Key-Value-Pair
(KVP) data model and a different number of technologies. Similarly,
\cite{cloud-comp-survey} reports similar trends in academic and industrial
research in distributed computing have used KVP databases in powerful systems
such as Cloud Computing \cite{cloud-comp-architectures}. For this reason, in
addition to the different database systems used by projects in the
literature review, this work also includes the review of a KVP database system
called mongoDB \cite{mongodb}, since it was among those cited in the surveys.
In such a manner, the list of the systems considered for assessment is the
following:

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  dateware house with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks
  implementation such as \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source KVP database
  system suggested by different surveys to be alternatives to relational
  databases \cite{db-is-rdbs-dommed, cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. Furthermore, it was selected because
  NetBEAMS represents serialized objects in XML, as shown in section
  \ref{sec:dsp-message}.
\end{itemize}

The following sections reviews these technologies' capabilities to adhere to
the specifications of the taxonomies defined in chapter 3, as well as the
requirements and the scope for a data persistence of NetBEAMS, as defined in
chapter 4.

@@

\begin{itemize}
  \item Data is generated by sensor devices, such as the YSI sonde
  \cite{YSI-Sonde}, and manually collected by using a laptop to the data sink
  at the RTC laboratories;
  \item Upon data reception, the RTC staff index and archive the data for
  distribution using the OPEnDAP format;
  \item An automated approach to the data collection procedures is performed 
  using NetBEAMS \cite{netbeams2009}, using software components to obtain and
  send the data to the data sink node. A complete description of this approach
  is detailed in section \ref{sec:netbeams-architecture}.
\end{itemize}

As summarized in the previous section, the nature of data from NetBEAMS through
SF-BEAMS is used for the purpose of \textbf{Data Archival}. After the
collected data captured from the sensors, it requires further processing
in order to be archived for later reuse. Different types of users are expected
to use the collected data, ranging from researchers, students, and the public 
over the Internet.

@@

Since all of developed database systems were implemented with the intension of
persisting data in long-term storage devices, every single one in the list
provides support for data persistence and management. Therefore, they are
compliant to the purpose of Data Archival for the sensor networks.

@@

As shown in the previous chapter, NetBEAMS's architecture is based on sensor
network whose topology is a single-hop star with a centralized data sink. In
this way, all the collected data is transmitted to that single location
\ref{sec:sn-infrastructure}. In the same vein, the simplest location to save
the collected data is through the use of at least an \textbf{External Storage}.
Similarly, the use of a Data-Centric Storage strategy is considered in
situations where the data sink is about to reach its storage capacity. As a
consequence, the \textbf{Data-Centric Storage} separates the collected data
into dissimilar locations, based on any property of the data. Therefore,
there are two different approaches to data partitioningThe two different
flavors:

\begin{itemize}
  \item \textbf{Horizontal Partitioning}: the data segmentation is implemented
  based on the values of the table rows. That is, all the values defined by the
  columns are used in a given table physically located in specific servers.
  For instance, horizontal partitioning can be defined as a collection of five
  years worth of historical sensor data, partitioned into five separate
  servers. In this way, each server contains data related to each of the years;
  \item \textbf{Vertical Partitioning}: data is partitioned vertically in a way
  that the columns of an entity are divided into different database tables. In
  this way, the particular dataset is broken down to different tables. For
  example, if the database is used to store pictures of a given sensor camera
  in a BLOB column, it would be the most preferable to the scenario where the
  images are not constantly referenced when compared to the textual data.
\end{itemize}

@@

The External Storage approach is the most common way to implement persistence
for any breed of systems since data management is usually offered by most
database systems out-of-the-box.

@@

The most common problem related to a single External Data Storage is that 
resources may run into data management problems such reaching the disk storage
capacity or hardware failure.

@@

The benefit of the Data-Centric approach can be directly associated with
database partitioning mechanisms described earlier. Since the collected data is
distributed into different locations, this approach decreases the size of
datasets when the database server performs a query processing in a database
table. For this reason, this approach helps the database administrators to
scale their database infrastructure and improve the database performance.

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-partitioning-relational}. Both the former and the
latter are used in regular schema-dependent models or schema-less models,
which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas, db-partitioning-relational}.

\begin{itemize}
  \item \textbf{Directory Based Partitioning}: an ordinary lookup service is
  used to identify which database shard to use, where the lookup table can be
  stored in a secondary area;
  \item \textbf{Key or Hash Based Partitioning}: the value of a given key
  property can  be used as the key of the shard. Moreover, it can be used as 
  an input to a hash function, whose output determines at which database shard
  the data needs to be stored.
\end{itemize}

An example of a Data-Centric storage with different shards can be see in Figure
\ref{fig:database-sharding-by-region}, where the collected data is segregated
by its origin. Each shard is assumed to be placed in different server
hosts, as well as each of them uses the same ``denormalized'' data models.

@@

The data partitioning is a highly restricted technology, normally available to
advanced users of distributed database systems. Despite being offered as a
common-off-the-shelf functionality by some popular database systems, this
strategy requires closer database management and configuration. Some of the
problems related to this strategy are regarding its lifecycle management
\cite{db-shard-discussion}:

\begin{itemize}
  \item \textbf{Rebalancing Shards}: considering a schema-dependent database
  system, changes on the schema of one shard requires rebalancing the
  collected data on each of the shards. As a consequence, the database shards
  transfers data to new locations whenever necessary. This technique is
  starting to appear as COTS implementations such as MySQL \cite{mysql} and
  mongoDB \cite{mongodb};
  \item \textbf{Referential integrity}: also considering a schema-dependent
  database system, any cross-shard query may fall into the problem where
  references do no exist in other shards, since the application layer is
  responsible for enforcing data integrity. In this way, examples of such
  include verifying constraints of foreign keys when the partition is done
  using Vertical table partitioning, considering that data collections are
  organized in separate spaces, and physically placed in different shard nodes
  \cite{db-partitioning-relational}.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports External or Data-Centric Storage approaches.
  When Data-Centric Storage is used, both vertical and horizontal partitions
  \cite{db-partitioning-relational} can be chosen, using different strategies
  to partition the data;
  \item \textbf{TinyDB}: Only supports the External Storage;
  \item \textbf{MongoDB}: Supports External or Data-Centric Storage. When the
  latter approach is used, the horizontal partition strategy is available with
  automatic shards \cite{db-mongo-partition};
  \item \textbf{DB2}: Supports External and Data-Centric storage. Only supports
  vertical table partition for the data-centric storage \cite{db-db2-partition}.
\end{itemize}

All the selected database systems support the External Storage method. However,
they all differ with the support to Data-Centric Storage support. Therefore,
MySQL and mongoDB are promising choices, with the difference of the use of the
data model.

\section{Analysis of the Data Model}

One of the most common practices in the area of database system is to use the
relational model to persist data, although the application of the system may
not fit to solve the problem. In fact, the main users of sensor networks may
not possess any expertise in database systems or data modeling, based on the
fact that they are from different science areas. \cite{sn-programming-language}
suggested the use of programming languages abstraction instead of regular
database systems for researchers. For instance, SF-BEAMS is managed by Marine
Biologists without expertise in Data Management, Modeling systems, and for
this reason, one of the requirements of the system is trying the use of
schema-less approaches. The \textbf{Key-Value-Pair Data Model} or the
\textbf{Document-Oriented Data Model} seems to be the simplest choices of the
models.

@@

\begin{itemize}
  \item Data schema that constantly changes require constant database
  normalization processes, changes to structure, database management, and all
  the code that uses the data. Schema-Dependent databases always require
  refactoring of the schema to accommodate changes to the entites' properties;
  \item The approach of Relational Model might not fit the nature of the
  collected data from sensor devices, since they tend to change over time. 
  Furthermore, the support to time-series or provenance are not better
  supported \cite{sn-provenance};
  \item Some research have suggested changes to the SQL clauses to better
  support time-series when collecting data using the relational model 
  \cite{sn-db-newop}.
\end{itemize}

Other projects have used the XML data models for persistence. It falls
into the same category as the Relational Model using XML documents \cite{xml},
since XML documents must fulfill the specifications defined by an XML Schema
\cite{xml-schema}. This model uses XPath \cite{xml-xpath} technology for
querying documents, although some hybrid \cite{db2} technologies may still use
SQL \cite{sql} for that matter. In this way, a materialized version of the
data from the defined previous schema can be seen in Figure
\ref{fig:persistence-example-relational}.

@@

Although this data model has been used in different types of applications,
developers have tried to use the Relational Model to design an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. As mentioned earlier, with the
emergence of Internet applications, the necessity of such a model that does
not require constant refactoring led developers to propose such manner using
the relational model as documented in technical articles such as
\cite{db-kvp-in-relational01} and \cite{db-kvp-in-relational02}. Based on
these articles, consider the relational model depicted in Figure
\ref{fig:KVP-on-Relational-Model} as a data model for persisting data to our
case study using the key-value strategy. It is clear that this approach
contains problems: repetition occurs, since the nature of
time-series data requires the use of a timestamp key for each property of the sensor. Therefore, this strategy is not
well-suited to provide persistence data from collected sensor networks.

@@

The growth of distributed systems and the Internet have enabled the development
of powerful database servers that can be organized in the context of
infrastructure and how to represent models. A powerful up-coming approach is
the use of database systems that do not use structured language for querying
its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, which is also known as Document-Oriented, Distributed Hash
Table. Such a model has the following properties:

@@

There are no records of the use of this data model with Sensor Networks.
Different variants of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents
have a dynamic structure, freely described without the use of a
schema validation mechanism as used in XML Schema. One example of such
approach is the use of the JSON data format \cite{json}. mongoDB implements the
following:

@@

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describes the data, depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model is defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships \cite{db-normalization};
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

@@

\begin{itemize}
  \item \textbf{MySQL}: supports the Relational Data Model;
  \item \textbf{TinyDB}: supports the Relational Data Model; 
  \item \textbf{MongoDB}: supports Key-Value Pair Data Model;
  \item \textbf{IBM DB2}: supports the Relational Model and the Structured
  Data models.
\end{itemize}

\section{Analysis of the Data Description}

\textbf{Data Description} is the methods used to describe the collected data
from sensors, specially those produced by real-time data streams. Since those
type of data does not include descriptions of its values, data provenance
offers guidelines to improve the description of the collected data.

An additiona to data provenance, annotations is another technique used to
better describe the collected data from sensor devices. Its main purposes is to
provide means of identifying observations, creating indexes based on simple
keywords. For an environmental sensor networks, this is a valuable type of value
used to better manage the collected data.

\subsection{Advantages of the Provenance-Aware Data}

The injection of provenance data into the collected data enriches description
of a given event, and is always useful to determine different aspects of the
data. As mentioned in chapter 3, data provenance helps identifying the nature
of the data by the identification of the properties of a given sensor device,
for example. Similarly, provenance adds spatio-time data in order to correlate
the observed data with time a given location and time of occurrence.

\subsection{Disadvantages of the Provenance-Aware Data}

Additional data being inserted together with the raw data collected from the
sensor devices. This additional data may contribute with any decreasing number
of disk storage for the variable defined. Therefore, the decreasing amount
of disk space may be one the reasons to not use annotations.

\subsection{Advantages of the Annotations}

Annotations are used to describe the collect data even more, since the
additional ``notes'' of the data are being stored together with the data. For
example, the use of tags to a given entity may help identifying and different
observations regarding the data collected from sensor devices.

\subsection{Disadvantages of the Annotations}

This type of data does not make part of the collected data, nor provenance
information.

All the database systems technology supports the use of Data Provenance or
Annotations. The Key-Value Pair data model has the advantage of not maintaining
a schema design, and therefore, can add any type of data, and therefore,
mongoDB is the lead in this taxonomy.

\section{Analysis of the Query Processing Mechanism}

@@

All the database systems technology support a centralized query system. In
addition to that, they can be distributed into different distributed nodes when
used in a partioned environement.

\section{Analysis of the System Organization}

@@

Each of the databases were scored according to its support to the different
types of taxonomies, as well as a different set of requirements. As defined
in the scope of the database system, the database that can provide a
programming language abstracton to its data access would be selected.
Moreover, the database technology should offer the support to schema-less
databases. As a result, mongoDB was selected because of different reasons.
First, it is a document-oriented database system, which does require the use of
SQL languages. Similarly, it had been reported in two different surveys as
potential database systems. Finally, there were no references in the literature
regarding the use o KVP databases in sensor networks, and for this reason, this
proposal was performed, supporting most of the taxonomies defined in the
previous chapter, it was selected for the implementation of the DSP Data
Persistence component for NetBEAMS.

\begin{table}
    \label{tab:technology-selection}
    \begin{center}
        \begin{tabular}[!h]{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{TinyDB} & \textbf{MongoDB} & \textbf{IBM DB2}\\\hline
        Centralized Query & + & + & + & + \\\hline 
        Distributed System & ++ & - & ++ & +\\\hline 
        Schema-less & - & - & + & -\\\hline 
        Provenance Support & + & + & + & +\\\hline 
        NO-SQL Query & - & - & + & -\\\hline 
        Partition & + & - & + & +\\\hline 
        Export Capability & + & - & + & -\\\hline 
        Programming Language. & + & + & + & +\\\hline
        Script Language. & + & - & + & +\\\hline
        Open-Source & + & + & + & -\\\hline
        \end{tabular}
        \caption{Technologies Selection Grading}
    \end{center}
\end{table}

To conclude, this chapter evaluated different database systems selected from
the literature reviews of chapter 2. After presenting the database system
``contenders'', the analysis of each of the taxonomies were conducted for each
of the technologies, culminating with the technology selection comparison. In
this way, the following chapter describes the design of the solution using
mongoDB as the primary solution for the backend.
